---
title: 'Supplementary Appendix B: Selected prior distributions'
output:
  word_document:
    reference_docx: "Appendix_template.docx"
---


# B.1 Pella-Tomlinson shape parameter

The Pella-Tomlinson surplus production model in terms of depletion $P_t = B_t / K$ where $B_t$ is biomass at time $t$ and $K$ is carrying capacity) is given by

$$P_t = P_{t-1} + P_{t-1}\frac{r}{m-1}\left(1 - P_{t-1}^{m-1}\right).$$

The parameter $m$ is known as the shape parameter. It is nonnegative, and encompasses other surplus production model at specific values, namely the Fox surplus production model at $m=1$ and the Schaefer surplus production model at $m=2$. The shape parameter determines the level of depletion where maximum sustainable yield occurs, $P_{MSY}$. The relationship between these is
$$P_{MSY} = g(m) = m^{-\frac{1}{m-1}}.$$

In order to parameterize a vaguely informative prior on $m$, the implied prior distribution on $P_{MSY}$ was examined. For a given probability density function [PDF] of the prior assigned to $m$, $p_m$, the PDF of $F_{MSY}$, $p_{PMSY}$, is given by
$$p_{PMSY}(P_{PMSY}) = \left\vert\frac{\textrm{d}}{\textrm{d}\,P_{MSY}} g^{-1}(P_{MSY}) \right\vert f_m(g^{-1}(P_{MSY}))$$
The inverse of the transformation function is given in terms of the Lambert $W$ function[^1],
$$g^{-1}(P_{MSY}) = \frac{W(P_{MSY} \log P_{MSY})}{\log P_{MSY}}.$$
Its first derivative is given by[^2]
$$\frac{\textrm{d}}{\textrm{d}\,P_{MSY}} g^{-1}(P_{MSY}) = \frac{\left(\log P_{MSY} - W(P_{MSY}\,\log P_{MSY})\right) W(P_{MSY}\,\log P_{MSY})}{P_{MSY}\,\log^2 P_{MSY} \left(W(P_{MSY}\,\log P_{MSY}) + 1\right)}.$$
In both the inverse function and its derivative, care must be taken to choose the correct branch of the Lambert $W$ function depending on whether $0 < P_{MSY} < \frac{1}{e}$ or $\frac{1}{e} < P_{MSY} < 1$.

[^1]: Wolfram|Alpha. Accessed 2019-08-31. <https://www.wolframalpha.com/input/?i=inverse+m%5E%28-1%2F%28m-1%29%29>
[^2]: Wolfram|Alpha. Accessed 2019-09-06. <https://www.wolframalpha.com/input/?i=d%2FdP+LambertW%28P+log%28P%29%29%2Flog%28P%29>

Given these, the PDF of $P_{MSY}$ can be computed given the form of the prior distribution on $m$ and that distribution's parameters. Here a skew-normal prior is placed on $\log(m)$. This was chosen to control the right skew in the prior on $P_{MSY}$. Right skew was preferred to reduce the prior density on lower values of $P_{MSY}$, which are associated with unreasonably large population growth rates at low population levels. A median $P_MSY$ of $0.4$ was chosen based on the literature (e.g. Winker et al., 2018[^3]).

[^3]: Winker, H., Carvalho, F., Kapur, M., 2018. JABBA: Just Another Bayesian Biomass Assessment. Fish. Res. 204, 275–288. https://doi.org/10.1016/j.fishres.2018.03.010)

The location ($\xi$), scale ($\omega$), and shape ($\alpha$) parameters of the log skew normal distribution were chosen to be $-1/2$, $1$, and $10$, respectively. The resulting PDFs can be plotted as

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
source("src/03_priorfunctions.R")
source("src/50_makefigs.R")
library(tidyverse)
library(rstan)

## Log-skew-normal prior on m
xi <- -0.5       # location
omega <- 1       # scale
alpha <- 10      # shape (skew)

n_dens <- 2049


## Calculate probability density of m
m <- seq(0.01, 10, len = n_dens)
dm <- dlsn(m, xi, omega, alpha)
## Calculate probability density of Pmsy
pmsy <- seq(0.001, 0.999, len = n_dens)
dpmsy <- dlsn_pmsy(pmsy, xi, omega, alpha)

dfd <- data.frame(par = rep(c("m", "Pmsy"), each = n_dens),
                  parval = c(m, pmsy),
                  dpar = c(dm, dpmsy))
```

```{r echo=FALSE, fig.width=8, fig.cap="Figure B.1: Log skew normal prior on the Pella-Tomlinson shape parameter ($m$, left), and implicit prior on $P_{MSY}$ (right)."}
m_plot <- dfd %>%
  filter(par == "m") %>%
  ggplot(aes(x = parval, y = dpar)) +
  geom_line() +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0, 10, 1),
                     expand = expand_scale()) +
  labs(title = "", x = expression(m), y = "") +
  theme_jkb(12)
  
pmsy_plot <- dfd %>%
  filter(par == "Pmsy") %>%
  ggplot(aes(x = parval, y = dpar)) +
  geom_line() +
  scale_y_continuous(expand = expand_scale(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(0, 1, 0.1),
                     expand = expand_scale(add = c(0, 1e-5))) +
  labs(title = "", x = expression(F[MSY]), y = "") +
  theme_jkb(12)

grid.arrange(m_plot, pmsy_plot, nrow = 1)
```

The quantiles of $m$ and $P_{MSY}$ under this prior are given in *Table B.1*. These quantiles cover a large range of feasible values and include the Fox surplus production model and the two surplus production specifications used here, with $F_{MSY} = 0.4$ and $F_{MSY} = 0.5$ (the Schaefer model).

```{r, echo=FALSE, cache=TRUE, fig.width=10}
n <- 1000000

## Use random draws to validate pdf calculations
rm <- exp(rsn(n, xi, omega, alpha))
rpmsy <- vapply(rm, m_to_pmsy, 1.0)

## Calculate quantiles
qtpmsy <- quantile(rpmsy, c(0.025, 0.05, 0.1, 0.25, 0.5,
                            0.75, 0.9, 0.95, 0.975))
qtm <- quantile(rm, c(0.025, 0.05, 0.1, 0.25, 0.5,
                      0.75, 0.9, 0.95, 0.975)) 
qt <- bind_rows(qtm, qtpmsy) %>%
  mutate(` ` = c("$m$", "$P_{MSY}$")) %>%
  select(` `, everything())
tab <- flextable::flextable(qt, theme_fun = flextable::theme_alafoli)
flextable::colformat_num(tab, col_keys = names(qtm), digits = 2)
# knitr::kable(qt, digits = 2, padding = 2, format = "markdown",
#              caption = "*Table B.1* Quantiles of the Pella-Tomlinson shape parameter $m$ and depletion at MSY, $P_{MSY}$.")
```

# B.2 Catch observation variance

For interpretability, a prior was placed on the coefficient of variation of the catch observation error $\xi^*$ rather than directly on the variance parameter $\xi$. Catch observation errors on the order of 1,000 tonnes seem reasonable, leading to considering a coefficient of variation of around 0.05 as the upper end of possible values. Motivated by penalized complexity priors[^4], an exponential prior was placed on $\xi^*$ with rate chosen so that $\Pr(\xi^* < 0.05) = 0.9$, so that
$$\xi^* \sim \operatorname{Exponential}(46).$$ This exponential prior also serves to shrink the model towards a simpler model with no error in catch observations.

[^4]: Simpson, D., Rue, H., Riebler, A., Martins, T.G., Sørbye, S.H., 2017. Penalising model component complexity: a principled, practical approach to constructing priors. Stat. Sci. 32, 1–28. https://doi.org/10.1214/16-STS576

# B.3 Instantaneous fishing mortality

With very little prior information on instantaneous fishing mortality for each year, an uninformative prior was specified based on a uniform prior on the proportion of biomass removed by fishing in year $t$, $\theta_t$, which is given by $$\theta_t = h^{-1}(F_t) = \exp(-F_t).$$ Given a prior density on $\theta$ $p_\theta$, the PDF of $F_t$ is $$p_{F}(F_t) = \left\lvert -\exp(-F_t) \right\rvert p_{\theta}(h^{-1}(F_t)).$$ For $$p_\theta(\theta_t) = 1 \quad t=1,\ldots,T,$$ this implies $$F_t \sim \operatorname{Exponential}(1)\quad t=1,\ldots,T.$$
